---
title: "Assignment 3: Instrumental Variable Estimation"
subtitle: "Replicating the IV strategy in Stokes (2015)"
author: "EDS 241 / ESM 244 (DUE: 2/17/2026)"
format:
  html:
    theme: sketchy
    css: styles.css
date: "February 10, 2026"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

------------------------------------------------------------------------

### Assignment instructions:

Working with classmates to troubleshoot code and concepts is encouraged. If you collaborate, list collaborators at the top of your submission.

All written responses must be written independently (in your own words).

Keep your work readable: Use clear headings and label plot elements thoughtfully (where applicable).

Submit both your rendered output and the Quarto (`.qmd`) file.

**Assignment submission (YOUR NAME)**: Zach Loo


------------------------------------------------------------------------

### Introduction

In this assignment, you will replicate the instrumental variable analysis from Stokes (2015), which examined how local wind turbine projects influenced electoral outcomes. Building on the matched dataset from Assignment 2, we will use Two-Stage Least Squares (2SLS) to estimate the causal effect of having a wind turbine proposed nearby on the change in Liberal Party vote share between 2007 and 2011. The instrument used in Stokes (2015) is a measure of local wind resource (average wind power, logged), which predicts where wind turbines are proposed. By using this instrument, we aim to isolate the portion of variation in turbine placement that is as-good-as-random, helping to meet the assumptions for causal identification.

**Study:** [Stokes, 2015 – Article](https://drive.google.com/file/d/1y2Okzjq2EA43AW5JzCvFS8ecLpeP6NKh/view?usp=drive_link)

**Data source:** [Dataverse – Stokes, 2015 replication data](https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/SDUGCC)

------------------------------------------------------------------------

::: callout
Note: The estimates you obtain may not exactly match the published results in Stokes (2015) due to the alternative matching procedure used for processing the data in the previous assignment. Estimates should approximate the findings reported in *Table 2* of the article.
:::

------------------------------------------------------------------------

### Load packages

```{r}
library(tidyverse)
library(janitor)
library(here)
library(jtools)   # for export_summs (pretty regression tables)
library(AER)      # for ivreg (2SLS estimation)

```

------------------------------------------------------------------------

### Load the matched dataset (from Assignment 2)

> The matched_data has been preprocessed by matching on key covariates (e.g. pretreatment home values, education, income, population density) to improve balance between treated and control precincts. We will now use this data for the IV analysis. Make sure to re-code the `precinct_id` variable as a `factor`. 

```{r}
matched_data <- read_csv(here::here("data", "matched_data_subset.csv")) |> 
  mutate(precinct_id = factor(precinct_id))

```

------------------------------------------------------------------------

### Part 1: IV Identification Rationale

Intuition for Using an Instrument: 

**Question 1:** After matching on observables, why might we still need to utilize an instrumental variable approach to identify the causal effect of turbine proposals on vote share? In other words, what potential issues remain that an IV method can help address in this context? Use specific examples from the study to illustrate threats to a causal interpretation, then explain how an IV approach is designed to mitigate those threats.

*Response:* Even after matching on observable characteristics, unobservable differences still remain. These unobserved factors could drive both turbine proposals and vote shares. Using wind speed which is determined by geography and climate is unrelated to local politics, thus this isolates the variation in turbine proposals that is exogenous to unobserved confounders.

------------------------------------------------------------------------

### Part 2: Two-Stage Least Squares (2SLS) Step-Wise Implementation

#### 2A. First-Stage Estimation: Regress the treatment ($D$) on the instrument ($Z$)

$$D_i = \alpha_0 + \alpha_1 Z_i $$

a. Estimate the first-stage regression of the treatment on the instrument (with controls). Regress `proposed_turbine_3km` on `log_wind_power`.
b. Include the control variables used in Stokes (2015) for both stages: Distance to lakes, geographic coordinates (latitude & longitude) with their squares and interaction, plus district fixed effects. 
c. After running the first stage, report the F-statistic for the instrument.

```{r}
first_stage <- lm(proposed_turbine_3km ~ log_wind_power + mindistlake + latitude + longitude + lat_sq + long_sq + latitude*longitude + factor(district_id), data = matched_data)

export_summs(first_stage, digits = 3,
             model.names = c("First stage: Proposed Turbine 3km"),
             coefs = c("(Intercept)", "log_wind_power")) 

```

###  Testing Instrument Relevance

Check instrument strength (F-statistic)
```{r}
first_stage_summary <- summary(first_stage)
first_stage_summary$fstatistic[1]
```

**Question 2A:** Based on the instrument relevance test reported in the study, would you conclude the instrument is strong enough to be credible? Explain what a weak instrument would mean in this setting: Specifically, what would it suggest about compliance with Ontario’s Green Energy Act policy? 

*Response:* The f-statistic of the instrument is reported to be 15.16, which is considered to be a strong instrument. A weak instrument would indicate that wind speed doesn't reliably predict where turbines are proposed. This suggests that wind developers are not proposing turbines based on wind resources but on other factors like political reception or local economic conditions.

------------------------------------------------------------------------

#### 2B. Second Stage Estimation 

#### Regress the outcome ($Y$) on the fitted values from the 1st stage ($\hat{X}_i$)

$$Y_i = \beta_0 + \beta_1 \hat{D}_i + \epsilon_i$$

a. Now estimate the second stage of the 2SLS. 
b. First, use the first-stage model to generate the predicted values of proposed_turbine_3km for each precinct (these are $\hat{D}_i$). 
c. Add these predicted values as a new column in matched_data (e.g. proposed_turbine_3km_HAT). 
d. Then, regress the outcome change_liberal (the change in Liberal vote share from 2007 to 2011) on the predicted treatment (proposed_turbine_3km_HAT), including the same controls and fixed effects as in the first stage. 
e. Fill in the code for these steps below to obtain the second-stage regression results.

Save predicted values $\hat{X}_i$ from first stage 
```{r}

matched_data$proposed_turbine_3km_HAT <- predict(first_stage)

```

#### Estimate the second-stage regression 

$$LiberalVoteShare_i = \beta_0 + \beta_1 \widehat{ProposedTurbine}_i + ControlVariables... +  \epsilon_i$$

```{r}
second_stage <- lm(change_liberal ~ proposed_turbine_3km_HAT + mindistlake + latitude + longitude + lat_sq + long_sq + latitude*longitude + factor(district_id), data = matched_data)

export_summs(second_stage, digits = 3, 
             model.names = c("Second stage: Change in Liberal Vote Share"),
             coefs = c("(Intercept)", "proposed_turbine_3km_HAT") ) 
```

#### Interpreting the 2SLS Estimate

**Question 2B:** Imagine you are explaining your 2SLS findings to a policymaker in Ontario. What does the estimated coefficient on `proposed_turbine_3km_HAT` imply about the electoral impact of a local wind turbine proposal (within 3 km) on liberal vote share?

*Response:* A wind turbine proposal causes a 10% drop in Liberal vote share. This supports the hypothesis that a backlash does exist on the political party that implemented the Green Energy Act.

**Question 2C:** Explain what it means that IV identifies a LATE in the context of the wind-turbine voting study. What specific subset of observations does the second-stage 2SLS estimate apply to, and what does it imply about interpretation and generalizability? 

*Response:* IV estimates the causal effect only for districts where turbine proposals were actually influenced by wind availability. In other words, the -0.102 effect only applies to districts that had turbine proposals determined by wind speed.

------------------------------------------------------------------------

### Part 3: IV Assumptions and Validity

#### Evaluate Instrument Validity 

**Question 3:** List the four key assumptions required for the IV strategy (2SLS) to identify a causal effect, and briefly explain what each one means in the context of this study. (Hint: think about what conditions a valid instrument must satisfy (relevance, exclusion,...)

*Response:* 1. Relevance of instrument, wind speed must strongly predict proposed turbines, this is satisfied in this study with an IV f-statistic > 10. 2. Independence of instrument, wind speed must not also affect unobserved confounders that affect voting, wind resources are affected by geography and climate, not from any local political or economic factors so this is also satisfied. 3. Exclusion of effect, wind speed affects voting only through turbine proposals and not directly. 4. No defiers, higher wind speeds must correlate to higher proposals in districts, not result in lower proposals somehow.

------------------------------------------------------------------------

### Part 4: Estimate 2SLS using `AER::ivreg()` 

📜 [SEE Documentation for specification details: AER package Viggnette Example](https://cran.r-project.org/web/packages/AER/vignettes/AER.pdf)

::: callout-tip

Syntax for specifying 2SLS using `ivreg()`:

`ivreg( Y ~ D + CONTROLS  | Z + CONTROLS , data )`

- The first-stage predictor variables go after the `~` symbol
- The second-stage predictor variables go after the `|` symbol

::: 


```{r}
  
fit_2sls <- ivreg(change_liberal ~ proposed_turbine_3km + mindistlake + latitude + longitude + lat_sq + long_sq + latitude*longitude + factor(district_id) | log_wind_power + mindistlake + latitude + longitude + lat_sq + long_sq + latitude*longitude + factor(district_id), data = matched_data)

export_summs(fit_2sls, digits = 3,
             model.names = c("Change in Liberal Vote Share"),
             coefs = c("(Intercept)", "proposed_turbine_3km")
             ) 
```

#### Robustness checking strategies utilized in Stokes, 2015

**Question 4:** Choose two *robustness checks* from the paper that the authors use to increase confidence in their causal identification strategy. For each one, summarize the logic and findings from the robustness check in your own words:

*Response:* One robustness check was the use of wind power as an instrumental variable. This was done because there was concerns about selection bias in turbine proposal locations, in other words, local political conditions influencing turbine placement. Wind power ended up being a strong instrument (F-stat = 68) and resulted in a second stage estimate of -0.077, meaning about an 8% decline in Liberal vote share for precincts within 3km of a turbine proposed. Second, the author also compared the federal level of Liberal vote share to the provincial level. This was done to test for informed voters, as in whether voters were actually punishing the provincial Liberal party or they were being indiscriminate between the Federal and provincial level. The effect against the Federal was small, around a 1% decline, indicating that voters were targeting the provincial government who were actually responsible for the wind policy.

------------------------------------------------------------------------

### END

------------------------------------------------------------------------
